---
title: "Discrete stable distributions"
author: "Will Townes"
date: "`r Sys.Date()`"
output: html_document
editor: source
---

```{r}
library(tidyverse)
theme_set(theme_bw())
source("discrete_stable.R")
```

## Case of $\alpha>2$

It's unclear whether discrete stable exists for $\alpha>2$. Let's try it out
and see if we end up with negative probabilities. The candidate PGF is
$$G(z)=\exp\left\{\delta(z-1)-\gamma(1-z)^\alpha\right\}$$
with the constraint that $\gamma>0$ if $\lceil\alpha\rceil$ is odd and $\gamma<0$ otherwise. The recursive formula for the PMF is
$$p_n = \frac{1}{n}\left[(\delta+\alpha\gamma)p_{n-1}+\gamma\sum_{j=2}^n j(-1)^{j-1}\binom{\alpha}{j}p_{n-j}\right]$$
Note that some terms may have negative signs, depending on the parity of
$\lceil\alpha\rceil$ and $n$. This means we can't just apply the usual 
log-sum-exp trick. We may assume $\delta+\alpha\gamma\geq 0$ since it is necessary for $p_1\geq 0$. 

Let $\alpha^\star=\lceil\alpha\rceil$. Regardless of its parity,
$\binom{\alpha}{j}>0$ for all $j\leq\alpha^\star$.

Case I: $\alpha^\star$ is odd. Then $\gamma>0$. If $j>\alpha^\star$,
then $\binom{\alpha}{j}<0$ for $j$ even and $\binom{\alpha}{j}>0$ for $j$ odd.
Since this alternating sign is canceled by the $(-1)^{j-1}$ term, all terms for 
$j>\alpha^\star$ are positive. Therefore, the negative terms are 
$j=2,4,\ldots,\lfloor\alpha\rfloor$. Note that if $\alpha\leq 2$ this means all terms are positive.

Case II: $\alpha^\star$ is even. Then $\gamma<0$. If $j>\alpha^\star$,
then $\binom{\alpha}{j}<0$ for $j$ odd and $\binom{\alpha}{j}>0$ for $j$ even.
Since this alternating sign is the opposite of the $(-1)^{j-1}$ term, all terms for 
$j>\alpha^\star$ are negative. But multiplying by the negative $\gamma$ makes 
them all positive. Therefore, the negative terms are 
$j=3,5,\ldots,\lfloor\alpha\rfloor$. Again, for $\alpha\leq 2$ all terms are positive.

```{r}
# alpha<-2.5
# gamma<-0.9
# delta<-5
probs<-dstable_pmf_naive(20,delta=5.0,gamma=.9,alpha=2.5)
barplot(probs,names.arg=0:(length(probs)-1))
sum(probs)
```

We will now implement the PMF on the log scale which should have better numerical stability. Since there are some negative weights (only for $\alpha>2$), we need a more 
[general version of logSumExp](https://mathstoshare.com/log-sum-exp-trick-with-negative-numbers) that handles the signs. Suppose we want to evaluate
$$\sum_{i=1}^n a_i - \sum_{j=1}^m b_j$$
where all the $a_i$ and $b_j$ terms are nonnegative (ie, the $b_j$ part is from the negative terms in the original sum, all put together). We only want to work
with $\log a_i$ and $\log b_j$ and we want the sum to also be evaluated on the
log scale. Obviously, the sum of the $a_i$ terms must be larger than the $b_j$
sum. If that's so, let $A=\log \sum_{i=1}^n \exp (\log a_i)$ (ie, the ordinary)
logSumExp function applied to the $\log a_i$ vector. Similarly let
$B=\log \sum_{j=1}^m \exp (\log b_j)$. This is logSumExp applied to the log 
absolute value of all the negative terms. Then,
$$\log \left(\sum_{i=1}^n a_i-\sum_{j=1}^m b_j\right) = \log(e^A-e^B) = A + \log\left(1-\exp\big(-(A-B)\big)\right)$$
The function of $A-B$ on the right side is the `log1mexp` function which
can be obtained from VGAM, based on a paper by [Maechler](https://cran.r-project.org/web/packages/Rmpfr/vignettes/log1mexp-note.pdf) (2012). Since it is so simple, we just re-implement it here to avoid the package dependency.

Check that both implementations return the same values

```{r}
probs1<-dstable_pmf_naive(20,delta=5.0,gamma=.9,alpha=2.5)
#barplot(probs1,names.arg=0:(length(probs1)-1))
probs2<-dstable_pmf(20,delta=5.0,gamma=0.9,alpha=2.5)
barplot(probs2,names.arg=0:(length(probs2)-1))
max(abs(probs1-probs2))
sum(probs2)
```

Check the $\alpha=2$ case is consistent with Hermite reference implementation

```{r}
alpha<-2; gamma<- -0.7
delta<-alpha^2*(-gamma)-0.5
ph0<-hermite::dhermite(0:20,delta+alpha*gamma,-gamma)
ph1<-dstable_pmf_naive(20,delta,gamma,alpha)
ph2<-dstable_pmf(20,delta,gamma,alpha)
max(abs(ph1-ph0)) #verify correctness
max(abs(ph2-ph1)) #verify consistency
barplot(ph2)
```

## Finding parameter limits

### Minimum delta

For a given $\alpha,\gamma$ combination, $\delta$ must be sufficiently large for the PMF to be valid.

```{r}
alpha<-2.9; gamma<-1; 
find_delta_limit(alpha,delta)
alpha<-seq(from=2.05,to=9.8,by=.067)
gamma<- (-1)^(ceiling(alpha)+1)
delta<-mapply(find_delta_limit,alpha,gamma,MoreArgs=list(n=500))
plot(alpha,delta,type="l",log="y")
```

```{r}
alpha<-expand.grid(c(.0001,seq(from=.05,to=.95,by=.05),.9999), 0:5)
alpha<-rowSums(alpha)
alpha<-sort(c(1,2,alpha))
#alpha<- seq(from = 1.01, to = 6.01, by = 0.05)
gsign<- (-1)^(ceiling(alpha)+1)
gsign[alpha==1]<- -1
gabs<-c(0.1,0.5,1.0)

f<-function(g) {
  data.frame(
    alpha = alpha,
    delta = mapply(find_delta_limit, alpha, g*gsign, MoreArgs=list(n=500)),
    gamma = g
  )
}

df <- do.call(rbind, lapply(gabs, f)) 
df$gamma <- factor(df$gamma, levels = rev(gabs))

# Sign bar: one tile per alpha, with edges at midpoints so the uneven alpha grid is covered
mid <- (head(alpha, -1) + tail(alpha, -1)) / 2
sign_df <- data.frame(
  xmin = c(alpha[1], mid),
  xmax = c(mid, alpha[length(alpha)]),
  sign = factor(gsign, levels = c(-1, 1))
)

# Put the bar just below the lowest curve
yr <- range(df$delta[is.finite(df$delta)])
h  <- 0.03 * diff(yr)
sign_df$ymin <- yr[1] - 1.5 * h
sign_df$ymax <- yr[1] - 0.5 * h

ggplot(df, aes(x = alpha, y = delta, colour = gamma)) + ylab("delta star") +
  geom_rect(data = sign_df,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = sign),
            inherit.aes = FALSE) +
  geom_line(linewidth = 1) +
  scale_colour_discrete(
    labels = parse(text = paste0("group('|', gamma, '|') == ", rev(gabs)))
  ) +
  scale_fill_manual(
    values = c("-1" = "grey75", "1" = "grey35"),
    labels = c(expression(gamma < 0), expression(gamma > 0))
  ) +
  guides(colour = guide_legend(order = 1), fill = guide_legend(order = 2)) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.02, 0.98),
    legend.justification = c(0, 1),
    legend.title = element_blank()
  )
ggsave("fig/delta_limit.pdf",width=6,height=4)
```

### Maximum absolute value of gamma

For a given alpha,delta combination, |gamma| must be kept suitably small for the PMF to be valid. Here we check this numerically for alpha>2. 

```{r}
alpha<-2.9; delta<-10; 
find_gamma_limit(alpha,delta)
alpha<-seq(from=2.05,to=9.8,by=.067)
gamma<-vapply(alpha,find_gamma_limit,1.0,delta=5.0)
sigma<-(gamma*sin(alpha*pi))^(1/alpha) #possible alternate parameterization
plot(alpha,gamma,type="l")
plot(alpha,abs(gamma),type="l",log="y")
plot(alpha,sigma,type="l")
#plot(alpha,gamma*sin(alpha*pi),type="l")
```

```{r}
alpha  <- seq(from = 2.05, to = 6, by = 0.05)
deltas <- c(9, 6, 3)

f<-function(d) {
  data.frame(
    alpha = alpha,
    gamma = vapply(alpha, find_gamma_limit, 1.0, delta = d, n=500),
    delta = d
  )
}

df <- do.call(rbind, lapply(deltas, f))
df$delta <- factor(df$delta, levels = deltas, labels = paste0("delta=", deltas))

#linetype = delta,
ggplot(df, aes(x = alpha, y = abs(gamma), colour=delta)) + 
  geom_line(linewidth = 1) +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.98, 0.98),
    legend.justification = c(1, 1),
    legend.title = element_blank()
  )
  #  scale_linetype_manual(values = 1:3) +
  #legend.background = element_rect(colour = "black")
ggsave("fig/gamma_limit.pdf",width=6,height=4)
```

## Visualizing the histograms with different choices of alpha

crude plot gives the basic idea
```{r}
n<-30
probs<-dstable_pmf(n,delta=10.0,gamma=3.4,alpha=2.5)#,log=TRUE)
barplot(probs,names.arg=0:n)
barplot(probs,names.arg=0:n, ylim=c(0,.015))
x<-1:1000
px<-dstable_pmf(max(x),delta=10.0,gamma=3.4,alpha=2.5)#,log=TRUE)
plot(x,px[-1],type="l",log="xy",lwd=2)
abline(0.6,-3.5,lty=3,col="red",lwd=2)
lines(x,dpois(x,10),lty=2,col="blue",lwd=2)
```

nicer version with ggplot

```{r}
df<-data.frame(delta=10, 
              alpha=c(0.5,0.9,1.0,1.1,1.5,2.0,2.1,2.5,2.9,3.1,3.5,3.9))
df$gamma<-vapply(df$alpha,find_gamma_limit,1.0,delta=df$delta[1],tol=1e-6)
df$gamma[df$alpha<1]<- 1
df$gamma[df$alpha==1]<- -1
sel<-df$alpha > 1 & df$alpha<=2
df$gamma[sel]<- 0.98*df$gamma[sel]
#sel<-df$alpha > 1 & df$alpha<=2
#df$gamma[df$alpha==2]<-0.95*df$gamma[df$alpha==2]

ann <- df %>% mutate(label=paste0("gamma=='", formatC(gamma, digits=3, format="fg", flag="#"), "'"))
xmax<-500
lp<-apply(df,1,function(r){dstable_pmf(xmax,delta=r[1],gamma=r[3],alpha=r[2],log=TRUE)})
rownames(lp)<-paste0("x=",0:xmax)
#df$alpha_lab<-paste0("alpha=='", formatC(df$alpha, digits=1, format="f"),"'")
df<-cbind(df,t(lp))
df<-df %>% pivot_longer(cols=starts_with("x="),names_to="x",names_prefix="x=",values_to="lp")
df$x<-as.integer(df$x)
df$px<-exp(df$lp)

#histogram up to x=30
df_ss<-df %>% filter(x<=30)
ggplot(df_ss) + geom_bar(stat="identity",aes(x=x,y=px))+facet_wrap(~alpha,scales="free_y",ncol=3,labeller=label_bquote(alpha == .(alpha))) + 
  geom_text(data = ann, aes(x = Inf, y = Inf, label = label),
  hjust = 1.1, vjust = 1.5, size = 3, parse = TRUE) + 
    ylab("P(X=x)") + theme(strip.text = element_text(hjust = 1))
ggsave("fig/discrete_stable_pmfs.pdf",width=6,height=6)
```

Log-log plots

```{r}
df1<-subset(df,x>0)
df1$log10px<-df1$lp*log10(exp(1)) #convert to log10 scale
#find intercept term of log-log tail
dfit<-subset(df1,x>200)
plaw_int<-dfit %>% group_by(alpha) %>% summarise(int=mean(log10px-log10(x)*(-(1+alpha))))
poi_pmf<-df1 %>% group_by(alpha) %>% mutate(log10poi=dpois(x,delta,log=TRUE)*log10(exp(1))) %>% filter(log10poi>=min(log10px))

#poi_pmf<-data.frame(x=1:xmax)
#poi_pmf$log10px<-dpois(poi_pmf$x,delta,log=TRUE)*log10(exp(1))

ggplot(df1) + geom_line(aes(x=x,y=log10px))+scale_x_log10()+
facet_wrap(~alpha,scales="free_y",ncol=3,labeller=label_bquote(alpha == .(alpha)))+ylab("log10 P(X=x)")+geom_abline(data=plaw_int,aes(intercept=int,slope=-(1+alpha)),linetype="dotted",color="red")+
geom_text(data = ann, aes(x = Inf, y = Inf, label = label),
  hjust = 1.1, vjust = 1.5, size = 3, parse = TRUE) + 
    xlab("x (log10 scale)") + theme(strip.text = element_text(hjust = 1))+
    geom_line(data=poi_pmf,aes(x=x,y=log10poi),colour="blue",linetype="dashed")
ggsave("fig/discrete_stable_loglog.pdf",width=6,height=6)
```

### Discrete self-decomposability

```{r}
n<-21
pd<-data.frame(alpha=rep(c(0.8,1,1.5),2), sigma=2)
pd$gamma<-sigma2gamma(pd$sigma,pd$alpha)
#pd$gamma<-round(pd$gamma)
pd$delta[1:3]<- -pd$alpha[1:3]*pd$gamma[1:3]+0.1
pd$delta[c(4,6)]<- -pd$gamma[c(4,6)]*pd$alpha[c(4,6)]^2
pd$delta[5]<- -2*pd$alpha[5]*pd$gamma[5]
#pd$delta<-ceiling(pd$delta)
# pd$gamma[1:3]<- (pd$delta[1:3]-0.1)/pd$alpha[1:3] #valid PMF but not self-decomp
# pd$gamma[c(4,6)]<-(pd$delta[c(4,6)]-0.1)/pd$alpha[c(4,6)]^2 #self-decomp
# pd$gamma[5]<- (pd$delta[5]-0.1)/(2*pd$alpha[5])
# pd$sigma<-gamma2sigma(pd$gamma,pd$alpha)
pd$self_decomp<-rep(c("not discretely self-decomposable","discretely self-decomposable"),each=3)
pd$alpha_txt<-paste0("alpha=",pd$alpha)
x<-0:(n-1)
pd2<- pd %>% group_by(alpha_txt,self_decomp) %>% 
  reframe(x=x, px=dstable_pmf(n-1,delta,gamma,alpha))


pd$delta_txt<-paste0("delta= ",signif(pd$delta,3))
pd$gamma_txt<-paste0("gamma= ",signif(pd$gamma,3))
ggplot(pd2)+geom_col(aes(x=x,y=px))+
  facet_grid(alpha_txt~self_decomp,scales="free_y")+
  geom_text(data=pd,aes(x=Inf,y=Inf,label=gamma_txt),hjust=1.05,vjust=1.3)+
  geom_text(data=pd,aes(x=Inf,y=Inf,label=delta_txt),hjust=1.05,vjust=2.6)+
  theme(panel.grid.minor = element_blank())+ylab("P(X=x)")
#+geom_vline(aes(xintercept=delta))
ggsave("fig/self_decomp.pdf",width=6,height=6)

#transposed
pd3<-pd2 %>% mutate(
  self_decomp=fct_recode(self_decomp,
                         "not disc. self-decomp."="not discretely self-decomposable",
                         "disc. self-decomp."="discretely self-decomposable"))
pd3$self_decomp<-factor(pd3$self_decomp)
pd4<-pd %>% mutate(
  self_decomp=fct_recode(self_decomp,
                         "not disc. self-decomp."="not discretely self-decomposable",
                         "disc. self-decomp."="discretely self-decomposable"))
pd4$self_decomp<-factor(pd4$self_decomp)
ggplot(pd3)+geom_col(aes(x=x,y=px))+
  facet_grid(self_decomp~alpha_txt,scales="free_y")+
  geom_text(data=pd4,aes(x=Inf,y=Inf,label=gamma_txt),hjust=1.05,vjust=1.3)+
  geom_text(data=pd4,aes(x=Inf,y=Inf,label=delta_txt),hjust=1.05,vjust=2.6)+
  theme(panel.grid.minor = element_blank())+ylab("P(X=x)")
ggsave("fig/self_decomp_wide.pdf",width=6,height=4)
```
