---
title: "Discrete stable distributions"
author: "Will Townes"
date: "`r Sys.Date()`"
output: html_document
editor: source
---

```{r}
library(tidyverse)
theme_set(theme_bw())
#source("util.R")
```

## Case of $\alpha>2$

It's unclear whether discrete stable exists for $\alpha>2$. Let's try it out
and see if we end up with negative probabilities. The candidate PGF is
$$G(z)=\exp\left\{\delta(z-1)-\gamma(1-z)^\alpha\right\}$$
with the constraint that $\gamma>0$ if $\lceil\alpha\rceil$ is odd and $\gamma<0$ otherwise. The recursive formula for the PMF is
$$p_n = \frac{1}{n}\left[(\delta+\alpha\gamma)p_{n-1}+\gamma\sum_{j=2}^n j(-1)^{j-1}\binom{\alpha}{j}p_{n-j}\right]$$
Note that some terms may have negative signs, depending on the parity of
$\lceil\alpha\rceil$ and $n$. This means we can't just apply the usual 
log-sum-exp trick. We may assume $\delta+\alpha\gamma\geq 0$ since it is necessary for $p_1\geq 0$. 

Let $\alpha^\star=\lceil\alpha\rceil$. Regardless of its parity,
$\binom{\alpha}{j}>0$ for all $j\leq\alpha^\star$.

Case I: $\alpha^\star$ is odd. Then $\gamma>0$. If $j>\alpha^\star$,
then $\binom{\alpha}{j}<0$ for $j$ even and $\binom{\alpha}{j}>0$ for $j$ odd.
Since this alternating sign is canceled by the $(-1)^{j-1}$ term, all terms for 
$j>\alpha^\star$ are positive. Therefore, the negative terms are 
$j=2,4,\ldots,\lfloor\alpha\rfloor$. Note that if $\alpha\leq 2$ this means all terms are positive.

Case II: $\alpha^\star$ is even. Then $\gamma<0$. If $j>\alpha^\star$,
then $\binom{\alpha}{j}<0$ for $j$ odd and $\binom{\alpha}{j}>0$ for $j$ even.
Since this alternating sign is the opposite of the $(-1)^{j-1}$ term, all terms for 
$j>\alpha^\star$ are negative. But multiplying by the negative $\gamma$ makes 
them all positive. Therefore, the negative terms are 
$j=3,5,\ldots,\lfloor\alpha\rfloor$. Again, for $\alpha\leq 2$ all terms are positive.

```{r}
# alpha<-2.5
# gamma<-0.9
# delta<-5

validgamma<-function(gamma,alpha){
  if(alpha<=0){ #alpha must be positive
    return(FALSE)
  }
  if(gamma==0){ #if gamma=0, this is ordinary Poisson, so alpha=1 by convention
    return(alpha==1)
  }
  #hereafter assume gamma!=0
  gpos<- gamma>0
  if(alpha<1){
    return(gpos) #gamma should be positive
  } else if(alpha<=2){
    return(!gpos) #gamma should be negative
  } else { #alpha>2
    alpha_star<-ceiling(alpha)
    if(alpha==alpha_star){ #integer values of alpha>2 not allowed
      return(FALSE)
    } else if(alpha_star%%2==1){
      return(gpos) #gamma positive when ceil(alpha) is odd
    } else {
      return(!gpos)
    }
  }
}

check_delta<-function(delta,gamma,alpha){
  if(delta< -gamma*alpha){
    stop("Must have delta>= -gamma*alpha to ensure P(X=1) is nonnegative")
  } else if(delta< -gamma){
    stop("Must have delta>= -gamma to ensure P(X=0)<=1")
  }
  #if delta>= -alpha*gamma, all probabilities are valid for alpha<=2
  #if alpha>2, even if both of these conditions are met, it's still possible
  #to have invalid probabilities. These are caught dynamically by the 
  #PMF function since there doesn't appear to be a closed from constraint.
  #increasing delta or making |gamma| closer to zero will always
  #make the PMF more well behaved.
}

dstable_pmf_naive<-function(nmax,delta,gamma,alpha){
  #compute the discrete stable PMF for non-integer alpha
  #nmax is the largest nonnegative integer at which to evaluate the PMF
  #will return P(X=n) for n=0,1,...,nmax
  #delta is location parameter
  #gamma is scale parameter
  #alpha is shape (tail) parameter
  #all parameters can only be scalars
  #evaluates the PMF using a direct recursion method
  #may not have great numerical stability
  stopifnot(validgamma(gamma,alpha))
  #direct way (numerical underflow likely)
  #assume alpha is not an integer, for alpha=1,2 handle separately
  #nmax<-100
  #nvals<-0:nmax
  jvals<-2:nmax
  check_delta(delta,gamma,alpha)
  dag<-delta+alpha*gamma
  weights<-c(dag, gamma*jvals*choose(alpha,jvals)*(-1)^(jvals-1))
  probs<-vector("numeric",length=nmax+1)
  probs[1]<-exp(-delta-gamma) #P(X=0)
  if(nmax==0){ #edge case where only want P(X=0)
    return(probs[1])
  } 
  for(n in 1:nmax){
    probs[n+1]<- mean(probs[n:1]*weights[1:n]) #P(X=n), mean does 1/n part
    if(probs[n+1]<0){
      stop("negative probability encountered, increase delta or decrease absolute value of gamma")
    }
  }
  return(probs)
}
probs<-dstable_pmf_naive(20,delta=5.0,gamma=.9,alpha=2.5)
barplot(probs,names.arg=0:(length(probs)-1))
sum(probs)
```

We will now implement the PMF on the log scale which should have better numerical stability. Since there are some negative weights (only for $\alpha>2$), we need a more 
[general version of logSumExp](https://mathstoshare.com/log-sum-exp-trick-with-negative-numbers) that handles the signs. Suppose we want to evaluate
$$\sum_{i=1}^n a_i - \sum_{j=1}^m b_j$$
where all the $a_i$ and $b_j$ terms are nonnegative (ie, the $b_j$ part is from the negative terms in the original sum, all put together). We only want to work
with $\log a_i$ and $\log b_j$ and we want the sum to also be evaluated on the
log scale. Obviously, the sum of the $a_i$ terms must be larger than the $b_j$
sum. If that's so, let $A=\log \sum_{i=1}^n \exp (\log a_i)$ (ie, the ordinary)
logSumExp function applied to the $\log a_i$ vector. Similarly let
$B=\log \sum_{j=1}^m \exp (\log b_j)$. This is logSumExp applied to the log 
absolute value of all the negative terms. Then,
$$\log \left(\sum_{i=1}^n a_i-\sum_{j=1}^m b_j\right) = \log(e^A-e^B) = A + \log\left(1-\exp\big(-(A-B)\big)\right)$$
The function of $A-B$ on the right side is the `log1mexp` function which
can be obtained from VGAM, based on a paper by [Maechler](https://cran.r-project.org/web/packages/Rmpfr/vignettes/log1mexp-note.pdf) (2012). Since it is so simple, we just re-implement it here to avoid the package dependency.

```{r}
log1mexp<-function(x){
  #slightly modified from original VGAM implementation
  stopifnot(all(x[!is.na(x)]>0))
  ifelse(x<=log(2), log(-expm1(-x)), log1p(-exp(-x)))
}

signed_logSumExp<-function(lx,nn){
  #lx is log|x| where x could be positive or negative
  #nn is a logical vector indicating which elements of x are nonnegative
  #ie, nn<- sign(x)>=0
  #computes log(sum(x)) with numerical stability
  #we require that the sum of x be positive otherwise it throws an error
  #the sum is never directly computed though
  #lx is allowed to be -Inf (ie, x can have exact zeros)
  A<-matrixStats::logSumExp(lx,nn)
  B<-matrixStats::logSumExp(lx,!nn)
  A+log1mexp(A-B)
}

dstable_pmf<-function(nmax,delta,gamma,alpha,log=FALSE){
  #compute the discrete stable PMF for non-integer alpha
  #nmax is the largest nonnegative integer at which to evaluate the PMF
  #will return P(X=n) for n=0,1,...,nmax
  #delta is location parameter
  #gamma is scale parameter
  #alpha is shape (tail) parameter
  #all parameters can only be scalars
  stopifnot(validgamma(gamma,alpha))
  #assume alpha is not an integer, for alpha=1,2 handle separately
  jvals<-2:nmax
  check_delta(delta,gamma,alpha)
  dag<-delta+alpha*gamma
  ldag<-log(dag)
  #weights<-c(dag, gamma*jvals*choose(alpha,jvals)*(-1)^(jvals-1))
  lwts<-c(ldag, log(abs(gamma))+log(jvals)+lchoose(alpha,jvals))
  lp<-vector("numeric",length=nmax+1)
  lp[1]<- -delta-gamma #logP(X=0)
  if(nmax==0){ #edge case where only want P(X=0)
    return(ifelse(log,lp[1],exp(lp[1])))
  }
  nn<-rep.int(TRUE,nmax) #1,2,3,...,nmax. Which weights are nonnegative?
  #negative weights can only occur with alpha>2
  if(alpha>2){
    if(gamma>0){ #this implies ceil(alpha) is odd
      #negative terms are j=2,4,...floor(alpha)
      negterms<-seq(from=2,to=floor(alpha),by=2)
    } else { #case where ceil(alpha) is even
      #negative terms are j=3,5,...,floor(alpha)
      negterms<-seq(from=3,to=floor(alpha),by=2)
    }
    nn[negterms]<-FALSE
  }
  for(n in 1:nmax){
    terms<- lp[n:1] + lwts[1:n]
    lp[n+1]<- signed_logSumExp(terms,nn[1:n]) - log(n)
  }
  if(log){
    return(lp)
  } else {
    return(exp(lp))
  }
}
```

Check that both implementations return the same values

```{r}
probs1<-dstable_pmf_naive(20,delta=5.0,gamma=.9,alpha=2.5)
#barplot(probs1,names.arg=0:(length(probs1)-1))
probs2<-dstable_pmf(20,delta=5.0,gamma=0.9,alpha=2.5)
barplot(probs2,names.arg=0:(length(probs2)-1))
max(abs(probs1-probs2))
sum(probs2)
```

```{r}
find_gamma_limit_gt2<-function(alpha, delta, n=delta*10, tol=1e-3, gamma_start = 1e-3, gamma_max = 1e6){
  stopifnot(alpha > 2, alpha != round(alpha), delta >= 0, tol > 0)
  # sign convention from ceiling(alpha) parity
  gamma_sign <- if (ceiling(alpha) %% 2 == 1) 1 else -1
  is_valid <- function(g){
    tryCatch({
      dstable_pmf(n=n,delta=delta,gamma=g,alpha=alpha)
      return(TRUE)
    }, error = function(e) FALSE)
  }
  # anchor: |gamma| = 0 treated as valid lower bound
  lo <- 0
  hi <- gamma_start
  # expand hi geometrically until we find an invalid magnitude
  repeat {
    if (!is_valid(gamma_sign * hi)) break
    lo <- hi
    hi <- hi * 2
    if (hi > gamma_max) {
      stop("No invalid gamma found up to gamma_max = ", gamma_max,
           "; increase gamma_max or check pmf_fun.")
    }
  }
  # bisection: lo is valid, hi is invalid, shrink until width < epsilon
  while ((hi - lo) > tol) {
    mid <- (lo + hi) / 2
    if (is_valid(gamma_sign * mid)) {
      lo <- mid
    } else {
      hi <- mid
    }
  }
  return(gamma_sign*lo)
}

find_gamma_limit<-function(alpha,delta,...){
  #for a given value of tail index (alpha) and location (delta), return the gamma value with largest absolute value
  #such that the parameter combination is still valid but it is not valid for any gamma with larger abs val. 
  #based on numerical pmf evaluation of P(X=0),P(X=1),...,P(X=n)
  #... additional args passed to find_gamma_boundary_gt2
  stopifnot(alpha>0)
  if(alpha<1){
    #mean is not finite, gamma>0 is only lower bounded by -delta/alpha in the weird case of delta<0
    return(Inf)
  } 
  stopifnot(delta>=0)
  if(alpha<=2){ #gamma<=0 for alpha in [1,2]
    return(-alpha*delta)
  }
  find_gamma_limit_gt2(alpha,delta,...)
}

alpha<-2.9; delta<-10; 
find_gamma_limit(alpha,delta)
```


## Visualizing the histograms with different choices of alpha


### Alpha in (2,3)

```{r}
n<-30
probs<-dstable_pmf(n,delta=10.0,gamma=3.4,alpha=2.5)#,log=TRUE)
barplot(probs,names.arg=0:n)
barplot(probs,names.arg=0:n, ylim=c(0,.015))
x<-1:1000
px<-dstable_pmf(max(x),delta=10.0,gamma=3.4,alpha=2.5)#,log=TRUE)
plot(x,px[-1],type="l",log="xy",lwd=2)
abline(0.6,-3.5,lty=3,col="red",lwd=2)
lines(x,dpois(x,10),lty=2,col="blue",lwd=2)
```

```{r}
n<-30
probs<-dstable_pmf(n,delta=10.0,gamma=2.7,alpha=2.9)#,log=TRUE)
barplot(probs,names.arg=0:n)
barplot(probs,names.arg=0:n, ylim=c(0,.001))
x<-1:1000
px<-dstable_pmf(max(x),delta=10.0,gamma=2.7,alpha=2.9)#,log=TRUE)
plot(x,px[-1],type="l",log="xy",lwd=2)
abline(0.2,-3.9,lty=3,col="red",lwd=2)
lines(x,dpois(x,10),lty=2,col="blue",lwd=2)
```

make a pretty ggplot

```{r}
# scen<-data.frame(n=30,delta=10.0,gamma=c(3.4,2.7))
# pfunc<-function(
```

### Alpha in (3,4)

```{r}
n<-35
probs<-dstable_pmf(n,delta=10.0,gamma=-0.95,alpha=3.1)#,log=TRUE)
barplot(probs,names.arg=0:n)
barplot(probs,names.arg=0:n,ylim=c(0,1e-4))
x<-1:1000
px<-dstable_pmf(max(x),delta=10.0,gamma=-0.95,alpha=3.1)#,log=TRUE)
plot(x,px[-1],type="l",log="xy",lwd=2)
abline(-0.2,-4.1,lty=3,col="red",lwd=2)
lines(x,dpois(x,10),lty=2,col="blue",lwd=2)
```

```{r}
n<-35
probs<-dstable_pmf(n,delta=10.0,gamma=-1.5,alpha=3.25)#,log=TRUE)
barplot(probs,names.arg=0:n)
barplot(probs,names.arg=0:n,ylim=c(0,2e-4))
x<-1:1000
px<-dstable_pmf(max(x),delta=10.0,gamma=-1.5,alpha=3.25)#,log=TRUE)
plot(x,px[-1],type="l",log="xy",lwd=2)
abline(0.45,-4.25,lty=3,col="red",lwd=2)
lines(x,dpois(x,10),lty=2,col="blue",lwd=2)
```


```{r}
n<-50
probs<-dstable_pmf(n,delta=10.0,gamma=-1.22,alpha=3.5)#,log=TRUE)
barplot(probs,names.arg=0:n)
barplot(probs,names.arg=0:n,ylim=c(0,1e-4))
x<-1:1000
px<-dstable_pmf(max(x),delta=10.0,gamma=-1.22,alpha=3.5)#,log=TRUE)
plot(x,px[-1],type="l",log="xy",lwd=2)
abline(0.7,-4.5,lty=3,col="red",lwd=2)
lines(x,dpois(x,10),lty=2,col="blue",lwd=2)
```

```{r}
n<-50
probs<-dstable_pmf(n,delta=10.0,gamma=-0.81,alpha=3.9)#,log=TRUE)
barplot(probs,names.arg=0:n)
barplot(probs,names.arg=0:n,ylim=c(0,1e-5))
x<-1:1000
px<-dstable_pmf(max(x),delta=10.0,gamma=-0.81,alpha=3.9)#,log=TRUE)
plot(x,px[-1],type="l",log="xy",lwd=2)
abline(0.2,-4.9,lty=3,col="red",lwd=2)
lines(x,dpois(x,10),lty=2,col="blue",lwd=2)
```

### Alpha in (0,2]

Make a bunch of histograms

```{r}
# n<-21
# pd<-data.frame(alpha=rep(c(0.8,1,1.5),2), sigma=2)
# pd$gamma<-sigma2gamma(pd$sigma,pd$alpha)
# #pd$gamma<-round(pd$gamma)
# pd$delta[1:3]<-pd$alpha[1:3]*pd$gamma[1:3]+0.1
# pd$delta[c(4,6)]<-pd$gamma[c(4,6)]*pd$alpha[c(4,6)]^2
# pd$delta[5]<-2*pd$alpha[5]*pd$gamma[5]
# #pd$delta<-ceiling(pd$delta)
# # pd$gamma[1:3]<- (pd$delta[1:3]-0.1)/pd$alpha[1:3] #valid PMF but not self-decomp
# # pd$gamma[c(4,6)]<-(pd$delta[c(4,6)]-0.1)/pd$alpha[c(4,6)]^2 #self-decomp
# # pd$gamma[5]<- (pd$delta[5]-0.1)/(2*pd$alpha[5])
# # pd$sigma<-gamma2sigma(pd$gamma,pd$alpha)
# pd$self_decomp<-rep(c("not discretely self-decomposable","discretely self-decomposable"),each=3)
# pd$alpha_txt<-paste0("alpha=",pd$alpha)
# x<-0:(n-1)
# pd2<- pd %>% group_by(alpha_txt,self_decomp) %>% 
#   reframe(x=x, px=dpoistable(x,alpha,sigma,delta))

# pd$delta_txt<-paste0("delta= ",signif(pd$delta,3))
# pd$gamma_txt<-paste0("gamma= ",signif(pd$gamma,3))
# ggplot(pd2)+geom_col(aes(x=x,y=px))+
#   facet_grid(alpha_txt~self_decomp,scales="free_y")+
#   geom_text(data=pd,aes(x=Inf,y=Inf,label=gamma_txt),hjust=1.05,vjust=1.3)+
#   geom_text(data=pd,aes(x=Inf,y=Inf,label=delta_txt),hjust=1.05,vjust=2.6)+
#   theme(panel.grid.minor = element_blank())+ylab("P(X=x)")
# #+geom_vline(aes(xintercept=delta))
# ggsave("discrete_stable_pmfs.pdf",width=6,height=6)

# #transposed
# pd3<-pd2 %>% mutate(
#   self_decomp=fct_recode(self_decomp,
#                          "not disc. self-decomp."="not discretely self-decomposable",
#                          "disc. self-decomp."="discretely self-decomposable"))
# pd3$self_decomp<-factor(pd3$self_decomp)
# pd4<-pd %>% mutate(
#   self_decomp=fct_recode(self_decomp,
#                          "not disc. self-decomp."="not discretely self-decomposable",
#                          "disc. self-decomp."="discretely self-decomposable"))
# pd4$self_decomp<-factor(pd4$self_decomp)
# ggplot(pd3)+geom_col(aes(x=x,y=px))+
#   facet_grid(self_decomp~alpha_txt,scales="free_y")+
#   geom_text(data=pd4,aes(x=Inf,y=Inf,label=gamma_txt),hjust=1.05,vjust=1.3)+
#   geom_text(data=pd4,aes(x=Inf,y=Inf,label=delta_txt),hjust=1.05,vjust=2.6)+
#   theme(panel.grid.minor = element_blank())+ylab("P(X=x)")
# ggsave("discrete_stable_pmfs_wide.pdf",width=6,height=4)
```